---
title: "EX3_ANSWERS_RU_tidyverse_dplyr_group_and_summarize"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 3**  
<p>&nbsp;</p>
1. Group df1 by the variable _age_class_. 
2. Filter to get the biggest 5 by the variable _length_ in each group. 
3. Summarise this data frame over the variable _length_ by calculating the mean. 

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer for Q1
```{r}
# Group df1 by the variable _age_class_. 
df1_A <- group_by(df1, age_classbylength)
df1_A
```
## Answer for Q2
```{r}
# Filter to get the biggest 5 by the variable _length_ in each group. 
df1_A <- filter(df1_A, rank(-length_mm) <= 5)
df1_A
```
## Answer for Q3
```{r}
# Summarise this data frame over the variable _length_ by calculating the mean.
summarise(df1_A, mean_length_mm = mean(length_mm, na.rm = T))
```

