---
title: "EX4_ANSWERS_RU_tidyverse_pipes"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 4**  
<p>&nbsp;</p>
1. Filter to only look at _yearling_ in variable age_classbylength
2. Group by common_name
3. Create new variable that is z-score of _length_ across groups
4. Create boxplot of grouped length zscores

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer for Q1
```{r}
df1 %>% 
  filter(age_classbylength == 'yearling')
```
## Answer for Q2
```{r}
df1 %>% 
  filter(age_classbylength == 'yearling') %>%
  group_by(common_name)
```
## Answer for Q3
```{r}
df1 %>% 
  filter(age_classbylength == 'yearling') %>%
  group_by(common_name) %>% 
  mutate(length_zscore = scale(length_mm))
```
## Answer for Q4
```{r}
df1 %>% 
  filter(age_classbylength == 'yearling') %>%
  group_by(common_name) %>% 
  mutate(length_zscore = scale(length_mm)) %>%
  ggplot(aes(x = common_name, y = length_zscore)) +
  geom_boxplot(fill=c('gold','darkorange'))
```

