---
title: "EX7_ANSWERS_RU_tidyverse_dplyr_join"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 7**  
<p>&nbsp;</p>
1. Calculate CPMs for our expressed genes (tidy_counts_expressed)
2. Calculate TPMs for our expressed genes
3. Draw a plot to compare the two


Hint:

**Counts per million (CPM)** are the gene counts normalized to total counts in a sample, multiplied by a million to give you a sensible number.  

gene_A_CPM = (gene_A_counts / sum(all_genes_counts)) * 1,000,000

**Transcripts per million (TPM)** are the gene counts normalized to total counts in a sample, multiplied by a million to give you a sensible number.  

gene_A_TPM = (gene_A_counts / sum(all_genes_counts / all_genes_lengths)) * 1/gene_A_length * 1,000,000

More info on RNAseq counts quantification here:
http://luisvalesilva.com/datasimple/rna-seq_units.html

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer 1
```{r}
# Use Group to focus computation on each sample. Use mutate to make 
# new variable that is the CPM
tidy_counts_expressed_norm <- tidy_counts_expressed  %>% 
  group_by(Sample) %>% 
  mutate(CPM=(counts/sum(counts))*1000000)
```
## Answer 2
```{r}
# Join our tidy data to the metadata, then make new variable that is the TPM
tidy_counts_expressed_norm <- tidy_counts_expressed_norm %>% 
  inner_join(counts_metadata, by = c("ENTREZ" = "ID")) %>%  
  mutate(TPM=(counts/sum(counts/LENGTH))*(1000000/LENGTH))
```
## Answer 3
```{r}
# Simple X-Y plot comparing TPM and CPM. 
p <- tidy_counts_expressed_norm %>% 
  ggplot(aes(x=CPM, y=TPM)) + 
  geom_point() + 
  scale_x_continuous(name="log2(CPM)",trans='log2') + 
  scale_y_continuous(name="log2(TPM)",trans='log2')
p
```

