---
title: "EX9_ANSWERS_RU_tidyverse_stringr"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 9**  
<p>&nbsp;</p>
1. How many Hox genes (starting with HOX), or ATPase genes (starting with ATP), in our expressed genes list (tidy_counts_expressed_norm)?
2. Subset dataframe to just ATPase genes. 
3. Create a new lowercase variable that has 'ATP' removed from the symbol.

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer 1
```{r}
tidy_counts_expressed_norm %>% 
  mutate(ATPase = str_starts(SYMBOL,'ATP')) %>% 
  split(.$Sample) %>% 
  map(~sum(.$ATPase))

tidy_counts_expressed_norm %>% 
  mutate(Hox = str_starts(SYMBOL,'HOX')) %>%
  split(.$Sample) %>% 
  map(~sum(.$HOX))
```
## Answer 2
```{r}
tidy_counts_expressed_norm %>% 
  filter(str_starts(SYMBOL,'ATP'))
```
## Answer 3
```{r}
tidy_counts_expressed_norm %>% 
  filter(str_starts(SYMBOL,'ATP')) %>%
  mutate(ATPtype = str_replace_all(SYMBOL,'ATP','' )) %>%
  mutate(ATPtype = str_to_lower(ATPtype))

```

