---
title: "20191009_gettingtidy"
author: "Matt Paul"
date: "10/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BRC intro



## What does it mean to be tidy?

Every dataset is different. But if we are ointeracintg with them having a standarad format i.e. count matrix or

There are many ways to do things. Everyone has their favourite syntax. 

Many of us taught BaseR. Which is great. 

Although many fundamental data processing functions exist in R, they have been a bit convoluted to date and have lacked consistent coding and the ability to easily flow together → leads to difficult-to-read nested functions and/or choppy code.

Tidyverse is a collection of tools and a philosphy to approaching data analysis that often makes data wrangling easier. They're self described as an opionionated group of packages. But packages in the tidyverse community readily talk to one another. 

More efficient code
Easier to remember syntax
Easier to read syntax

## What are the main tools in the tidyverse?

readr – reading data into R
tidyr – creating tidy data frames (Tibbles)
dplyr – mainpulating data
ggplot2 – making pretty graphs
stringr – working with strings
purr - iterating over data
tibble - working with tibbles

Other tools have now been made for the tidyverse. Overlap with biocoudctor. But these are the linchpins that hold it together. 

## What are we doing today

Workflow Image for working with data. 

## First step lets load in the data we are using today
```{r}
load(file='~/Documents/Box Sync/RU/Teaching/teaching/tidyR/my_tidy.Rdata')
```

## Are all data frames equal?
```{r}
df1
df2
df3a
df3b
```

## What it is to be tidy?

A tidy dataset is a data frame (or table) for which the following are true:

* Each variable has its own column
* Each observation has its own row
* Each value has its own cell

Our first dataframe is tidy

## Why bother?

Consistent dataframe layouts help to ensure not just the values are present, but also the relationships between datapoints.

Tidy data works well with R because R is a vectorized programming language. Data structures in R are built from vectors and R’s operations are optimized to work with vectors. Tidy data takes advantage of both of these traits.

R stores tabular data as a data frame, a list of atomic vectors arranged to look like a table. Each column in the table is an atomic vector in the list.


## When its in this format it is then super easy to interact with -> dplyr

```{r}
library(dplyr)

#These tools hav same logic

# -> insert graphic here
# https://github.com/trinker/tidyr_in_a_nutshell

#Select allows you to make a vector or dataframe from a specific variable or variables
select(df1,common_name)
select(df1,age_classbylength,common_name)
select(df1, -length_mm)
select(df1, common_name:length_mm)

#Filter allows you to remove observations based on specific criteria
filter(df1,common_name=='Sockeye salmon')
filter(df1,common_name %in% c('Sockeye salmon', 'Sockeye salmon'))
filter(df1, str_ends(common_name, 'salmon'))
filter(df1,length_mm>200 | length_mm<120)

#Arrange sorts the dataframe based on a specific variable or variables
arrange(df1, length_mm)
arrange(df1, common_name, desc(length_mm))

#Mutate creates a new variable based on some form of computation
mutate(df1,scale(IGF1_ng_ml, center = TRUE, scale = TRUE))
mutate(df1,IGFngml_zscore=scale(IGF1_ng_ml, center = TRUE, scale = TRUE))

#Summarize applies aggregating or summary function to a group
df1_byname <- group_by(df1, common_name)
summarise(df1_byname, count=n())
summarise(df1_byname, IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T))

#Grouping can help ask questions with other fucntions

#Smallest two from each species
filter(df1_byname, rank(length_mm) <= 2)

#Need at least 5 from each species
filter(df1_byname, n() > 5)

#Make new vriable using within group metrics
mutate(df1_byname,IGFngml_mm=scale(IGF1_ng_ml, center = TRUE, scale = TRUE))

```

Piping to put them together.
```{r}

# -> same graphic as before , but extend
# https://github.com/trinker/tidyr_in_a_nutshell

#Piping allows you to pass the result from one expression into another. 


# Without pipe
df1_byname <- group_by(df1, common_name)
summarise(df1_byname, IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T))

#With pipe
group_by(df1, common_name) %>% summarise(IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T))



group_by(df1, common_name) %>% filter(length_mm>200) %>% summarise(IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T))


mutate(df1,size=if_else(length_mm>200, 'big_fish', 'small_fish')) %>% group_by(common_name, size) %>% filter() %>% summarise(IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T))

mutate(df1,size=if_else(length_mm>200, 'big_fish', 'small_fish')) %>% group_by(common_name, size) %>% filter() %>% summarise(IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T)) %>% filter(common_name!='Coho salmon')  %>% filter(common_name!='Sockeye salmon')

```

## Say want to graph what is going on with these fish? age species. IGF per length. 
```{r}
library(ggplot2)

mutate(df1,size=if_else(length_mm>200, 'big_fish', 'small_fish')) %>% group_by(common_name, size) %>% filter() %>% summarise(IGF1_ng_ml_ave=mean(IGF1_ng_ml, na.rm=T)) %>% filter(common_name!='Coho salmon')  %>% filter(common_name!='Sockeye salmon') %>% ggplot( aes(x=IGF1_ng_ml_ave,group=interaction(common_name, size))) + geom_bar(stat="identity")

```

CHALLENGE
```{r}
GET A SPECIFIC GROUP AND PLOT IT. 
```

ANSWER
```{r}
GET A SPECIIFC GROUP AND PLOT IT. 
```


## So we blasted through what being tidy can give you. So how we do get tidy? Reading the data in and mkaing tibbles.
Now lets tidy some data. We know how to work a little bit with some data. 

ReadR:
read_csv(): comma separated (CSV) files
read_tsv(): tab separated files
read_delim(): general delimited files
read_fwf(): fixed width files
read_table(): tabular files where columns are separated by white-space.
read_log(): web log files

```{r}
library(tidyverse)

#Base gives you everything
untidy_counts_base <- read.csv("~/Documents/Box Sync/RU/Teaching/teaching/tidyR/dataset/hemato_rnaseq_counts.csv")

#readr gives you a tibble
untidy_counts <- read_csv("~/Documents/Box Sync/RU/Teaching/teaching/tidyR/dataset/hemato_rnaseq_counts.csv")
untidy_counts <- read_csv("~/Documents/Box Sync/RU/Teaching/teaching/tidyR/dataset/hemato_rnaseq_counts.csv", col_types = cols(
    ENTREZ = col_character(),
    CD34_1 = col_integer(),
    ORTHO_1 = col_integer(),
    CD34_2 = col_integer(),
    ORTHO_2 = col_integer()
  ))

```

#subsetting tibbles

```{r}
untidy_counts[,1]
untidy_counts[1,]
untidy_counts[1]

untidy_counts[[1]]
untidy_counts$ENTREZ
```

#Converting Tibbles - Backl and Forth
```{r}
as_tibble(untidy_counts_base)
untidy_counts_base<-as_tibble(untidy_counts_base) %>% mutate_at(vars(ENTREZ), as.character)
untidy_counts_base

as.data.frame(untidy_counts_base)
```

Make your own tibble. Lets grab some metadata 
```{r}
library(org.Hs.eg.db)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

keys<-untidy_counts$ENTREZ
symbols<-select(org.Hs.eg.db, keys=keys,columns="SYMBOL", keytype="ENTREZID")
chrs<-select(TxDb.Hsapiens.UCSC.hg19.knownGene, keys=keys, columns="TXCHROM", keytype="GENEID")

geneExons <- exonsBy(TxDb.Hsapiens.UCSC.hg19.knownGene,by="gene")[keys]
txsLength<-sapply(geneExons, function(x){x %>% reduce() %>% width() %>% sum()})

counts_metadata<-tibble(ID=symbols$ENTREZID,SYMBOL=symbols$SYMBOL,CHR=chrs$TXCHROM,LENGTH=txsLength)

```

What is wrong with this dataframe from a tidy viewpoint?

* Each variable has its own column
* Each observation has its own row
* Each value has its own cell

```{r}
untidy_counts
```

A siongle avirable with multiple columns

How do we get tidy? - Pivot tools (formerly known as gather/spread )
```{r}
tidier_counts<-pivot_longer(untidy_counts, cols=c(-ENTREZ), names_to = c("Sample"), values_to = "counts")
   
pivot_wider(tidier_counts, names_from = c("Sample"), values_from = "counts")
```

What next?
```{r}
tidier_counts
```

Multiple variables in a single column

How do we get tidy? - Cleaning up
```{r}
tidy_counts<-separate(tidier_counts, Sample, sep="_", into=c("CellType","Rep"), remove=FALSE)
#UNITE IS ALSO A THING


#rememebr you can alawys pipe this all together as well
untidy_counts %>% pivot_longer(cols=c(-ENTREZ), names_to = c("Sample"), values_to = "counts") %>% separate(Sample, sep="_", into=c("CellType","Rep"), remove=FALSE)
```

So now we have two realted tidy dataframes. Can we put them together?

You need a key. Often we want this to be unique i.e. ENTREZ ID. 

```{r}

inner_join(tidy_counts,counts_metadata, by = c("ENTREZ" = "ID"))

```

## There are many ways to join things

Inner Join 
* Keeps all obsevrations in x and y with matching keys

Outer Join
* A left join keeps all observations in x and those in y with matching keys.
* A right join keeps all observations in y and those in x with matching keys. 
* A full join keeps all observations in x and y

## Example - I only want to look at expressed genes

```{r}
expressed_genes<-tidy_counts %>% group_by(ENTREZ) %>% summarise(count_total=sum(counts)) %>% filter(count_total>0)

left_join(tidy_counts,expressed_genes, by = c("ENTREZ" = "ENTREZ")) %>% print(n=20)
tidy_counts_expressed<-right_join(tidy_counts,expressed_genes, by = c("ENTREZ" = "ENTREZ"))
tidy_counts_expressed  %>% print(n=20)
```


## Example - I only want to look at expressed genes
```{r}
Filtering joins

```

## Calcucate CPMs and TPMs

Show diagram of the math

```{r}
tidy_counts_expressed <- tidy_counts_expressed  %>% group_by(Sample) %>% mutate(CPM=(counts/sum(counts))*1000000)

tidy_counts_expressed <- tidy_counts_expressed %>% inner_join(counts_metadata, by = c("ENTREZ" = "ID")) %>%  mutate(TPM=(counts/(LENGTH/1000))/(sum(counts)/1000000))


tidy_counts_expressed %>% ggplot(aes(x=CPM, y=TPM)) +geom_point() + scale_x_continuous(name="log2(CPM)",trans='log2') + scale_y_continuous(name="log2(TPM)",trans='log2') 
```

writing out
```{r}
write_csv etc

```






# tidy history and ethos

Hadley Wickham (Chief Scientist at RStudio) has been the driving force behind the tidyverse. It now serves as the basis for many R packages and functions. 

There is a lot of support at https://www.tidyverse.org/

Hadley also wrote a paper about why tidy data is best: www.jstatsoft.org/v59/i10/paper.




## BRC outro
```{r}

```







