---
title: "EX10_ANSWERS_RU_tidyverse_forcats"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 10**  
<p>&nbsp;</p>
1. Create new categorical variable for the counts dataframe (tidy_counts_expressed_norm) based on whether a gene is short (0-3000), medium (0-6000) or long (6000+)
2. Ensure categorical variable is a factor which is ordered from long, medium then short
3. Draw a boxplot of TPMs comparing between our new groups

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer 1
```{r}
tidy_counts_expressed_norm %>%
  mutate(length_cat=if_else(LENGTH<3000, 'short', if_else(LENGTH>6000, 'long', 'medium')))
```
## Answer 2
```{r}
tidy_counts_expressed_norm %>%
  mutate(length_cat=if_else(LENGTH<3000, 'short', if_else(LENGTH>6000, 'long', 'medium'))) %>%
  mutate(length_cat=as_factor(length_cat)) %>%
  mutate(length_cat=fct_relevel(length_cat,levels=c('long', 'medium','short')))
```
## Answer 3
```{r}
tidy_counts_expressed_norm %>%
  mutate(length_cat=if_else(LENGTH<3000, 'short', if_else(LENGTH>6000, 'long', 'medium'))) %>%
  mutate(length_cat=as_factor(length_cat)) %>%
  mutate(length_cat=fct_relevel(length_cat,levels=c('long', 'medium','short')))  %>% 
  ggplot(aes(x=length_cat, y=TPM)) + 
  geom_boxplot() + 
  scale_y_continuous(trans='log2')

```

