---
title: "EX2_ANSWERS_RU_tidyverse_dplyr_arrange_and_mutate"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 2**  
<p>&nbsp;</p>
1. Create a dataframe with a new variable that contains the rank of the _length_ variable. Arrange this new data frame by _IGF_ (lowest to highest).
2. Create a dataframe with a new variable that is *IGF*/*length*. Arrange by this new variable (highest to lowest).

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer for Q1
```{r}
# Create a dataframe with a new variable that contains the rank 
# of the _length_ variable. Arrange this new data frame by 
# _IGF_ (lowest to highest).
df1_A <- mutate(df1, length_rank=rank(length_mm))
arrange(df1_A, IGF1_ng_ml)

```
## Answer for Q2
```{r}
# Create a dataframe with a new variable that is _IGF_/_length_. 
# Arrange by this new variable (highest to lowest).
df1_A <- mutate(df1, IGF_perlength=IGF1_ng_ml/length_mm)
arrange(df1_A, -IGF_perlength)
```




