---
title: "EX8_ANSWERS_RU_tidyverse_purrr"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse/"
output: 
  html_document:
    toc: true # table of contents true
    toc_float: yes
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---
<p>&nbsp;</p>
These exercises cover the sections of [Data wrangling with tidy](https://rockefelleruniversity.github.io/RU_tidyverse/). 

All files can be found in the "dataset" directory.
<p>&nbsp;</p>
**Exercise 8**  
<p>&nbsp;</p>
1. Create a nested dataframe of df1. Specifically only of the two largest groups by common_name, and nest by the common_name. 
2. Create a linear model between length_mm and IGF1_ng_ml using purrr over the nested dataframe. Add the results back to the dataframe
3. Create a new nested variable containing projections for length_mm based on your linear model (the function *predict* will be helpful here).

# ANSWERS

## Load in dataset and packages
```{r}
load(file='dataset/my_tidy.Rdata')
library(tidyverse)
```

## Answer 1
```{r}
df1 %>% 
  filter(common_name!='Coho salmon') %>% 
  filter(common_name!='Sockeye salmon') %>% 
  group_by(common_name) %>% 
  nest()
```
## Answer 2
```{r}
df1 %>% 
  filter(common_name!='Coho salmon') %>% 
  filter(common_name!='Sockeye salmon') %>% 
  group_by(common_name) %>% 
  nest() %>%
  mutate(my_model = map(data, ~lm(length_mm ~ IGF1_ng_ml, data = ., na.action = na.omit)))

```
## Answer 3
```{r}
df1_nest <- df1 %>% 
  filter(common_name!='Coho salmon') %>% 
  filter(common_name!='Sockeye salmon') %>% 
  group_by(common_name) %>% 
  nest() %>%
  mutate(my_model = map(data, ~lm(length_mm ~ IGF1_ng_ml, data = ., na.action = na.omit))) %>%
  mutate(predictions = map(my_model, predict)) 

df1_nest

df1_nest %>% pull(predictions)

```

